Liveness-Based Garbage Collection for Lazy Languages

### 2015-11-27 paper/paper.tex

%%           &\makecell{ $\rho, (\rho', x, e, \cred{\sigma'})\!:\!S,
%%   H, \kappa, \cred{\sigma}$  $\rightsquigarrow \rho', S, H[\rho'(x) :=
%%     \kappa], e, \cred{\sigma'}$ }   &  \sc{const}
%% \\
%% \hline
%%           & \makecell{$\rho, (\rho', z, e, \cred{\sigma'})\!:\!S, H, (\CONS~x~y), \cred{\sigma}$  $\rightsquigarrow
%%   \rho', S, H[\rho'(z) := (\rho(x),\rho(y))], e, \cred{\sigma'}$}     &  \sc{cons} \\
%% \hline
%% $H(\rho(x)) \mbox{ is } (l_1, l_2)$ & \makecell{$\rho, (\rho', z, e,
%%   \cred{\sigma'} )\!:\!S, H, (\CAR~x), \cred{\sigma}$  $
%%   \rightsquigarrow \rho', S, H[\rho'(z) := H(l_1)], e, \cred{\sigma'}$}      &
%% \sc{car-whnf} \\
%% \hline
%% $H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S,
%%   H, (\CAR~x), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, x,
%%   (\CAR~x), \cred{\sigma})\!:\!S, H, s, \cred{(\clazy \cup \acar)\sigma }$}      &
%% \sc{car-clo}
%% \\

%% \hline
%% $H(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$ & $\rho,\, S,\,  H,\,
%% (\CAR~x), \cred{\sigma} \rightsquigarrow \rho', \,(\rho, addr(\langle
%% s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S,\, H,\, s, \, \cred{\sigma}$      &
%% \sc{car-1-clo} \\

%% \hline
%% $H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
%%  & \makecell{$\rho, (\rho', z, e, \cred{\sigma'})\!:\!S, H,
%%   (+~x~y), \cred{\sigma}$   $\rightsquigarrow \rho', S, H[\rho'(z)
%%     := H(\rho'(x)) + H(\rho'(y))], e, \cred{\sigma'}$}      &
%% \sc{prim-whnf} \\
%% \hline
%% $H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S,
%%   H, (+~x~y), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, x,
%%   (+~x~y), \cred{\sigma})\!:\!S, H, s, \cred{\clazy\sigma}$}      &
%% \sc{prim-1-clo} \\
%% \hline
%% $H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho,
%%   S, H, (+~x~y), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, y,
%%   (+~x~y), \cred{\sigma})\!:\!S, H, s, \cred{\clazy\sigma}$}      &
%% \sc{prim-2-clo} \\
%% \hline
%% {$\mathit{g}~\mbox{defined as}$
%% $~(\DEFINE~(g~\myvec{y})~e_{\mathit{g}})$}  & \makecell{$\rho, S, H,
%%   (g~\myvec{x}), \cred{\sigma}$  $\rightsquigarrow [\myvec{y} \mapsto
%%     \rho(\myvec{x})], S, H, e_{\mathit{g}}, \cred{\sigma}$}      &
%% \sc{funcall} \\
%% \hline
%% $\ell$ is a new location& \makecell{$\rho, S, H, (\LET~x\leftarrow
%%   s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
%%   \ell]\rangle], e, \cred{\sigma}$} &
%% \sc{let} \\ 
%% \hline
%% $H(\rho(x)) \ne 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
%%   \cred{\sigma}$  $\rightsquigarrow \rho, S, H,  e_1, \cred{\sigma}$} & \sc{if-true} \\
%% \hline
%% $H(\rho(x)) = 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
%%   \cred{\sigma}$   $\rightsquigarrow
%% \rho, S, H,  e_2, \cred{\sigma}$} & \sc{if-false} \\
%% \hline
%% $H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
%%   (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma}$ $\rightsquigarrow
%% \rho', (\rho, x, (\SIF~x~e_1~e_2),  \cred{\sigma})\!:\!S, H, s,
%% \cred{\clazy\sigma}$}
%% &
%% \sc{if-clo} \\
%% \hline
%% {$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho,
%%   (\rho', z, e, \cred{\sigma'})\!:\!S, H, (\SRETURN~x), \cred{\sigma}$  $\rightsquigarrow \rho', S, H[\rho'(z) := v], e,
%%   \cred{\sigma'}$} &
%% \sc{return-whnf}\\
%% \hline
%% $H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
%%   (\psi:\SRETURN~x), \cred{\sigma}$  $
%%   \rightsquigarrow$
%% $\rho',~ (\rho, x, (\SRETURN~x), \cred{\sigma})\!:\!S, H,  s,
%%   \cred{\sigma}$} &
%% \sc{return-clo} \\
%% \hline
%% \end{tabular}
%% \caption{Minefield semantics.\label{fig:minefield-semantics}}
%% \end{center}
%% \vskip -5mm
%% \end{figure*}

We shall now present a proof of soundness of the analysis presented in
Section~\ref{sec:liveness-analysis}.  %% \warning{While  the proof  is on
  %% the broad lines of  \cite{asati14lgc}, there are several differences
  %% due to the  lazy semantics of the language in  this paper.}
Here are
the ideas behind the proof.
\begin{enumerate}
\item       We   augment     the        standard       semantics
  (Figure~\ref{fig:lang-semantics}) to model  a liveness-based garbage
  collection  just  before  each  $\rightsquigarrow$  transition.  The
  simulated garbage collection starts from  the root-set and inserts a
  special value  $\bot$ at each location  in the heap that  contains a
  reference that is reachable but not live. Any attempt to dereference
  such  locations during  the  transition will  result  in entering  a
  special state denoted \bang. We call the semantics after
  augmentation,  \emph{minefield semantics}.
\item \label{inline} Assuming that a  program enters the \bang\ state,
  we  show  how to  construct,  through  inline expansion,  a  program
  without function calls which has the same minefield behavior.
\item  The final  step   shows  that  no program  without
  function calls can enter a \bang\ state. As a consequence no program
  (with or without function calls) can enter the \bang\ state. 
\end{enumerate}

@@

\hline
\makecell[t]{$\cred{GC(\rho_1, S, H_1, (\CAR~x), \sigma) = (\rho, S, H)}$,} $H(\rho(x)) \mbox{ is } \bot$ & $\rho, S,
  H, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$   &
\sc{car-clo-bang} 
\\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, H_1, (\CAR~x), \sigma) = (\rho, S, H)}$,} $H(\rho(x))
\mbox{ is } \langle s, \rho'\rangle$ & $\rho, S,
  H, (\CAR~x), \cred{\sigma} \rightsquigarrow   \rho', (\rho, x,
  (\CAR~x), \cred{\sigma})\!:\!S, H, s, \cred{(\clazy \cup \acar)\sigma }$         &
\sc{car-clo}\\
\hline
\end{tabular}
\end{center}
\caption{Minefield semantics for \sc{car-clo-bang} and
  \sc{car-clo-bang} \label{fig:minefield-semantics-for-some}}
\end{figure*}

To set up the minefield semantics, we follow these steps:
\begin{enumerate}
\item   We   extend    the   abstract   machine   state
  $\rho,S,H,e$   to  $\rho,S,H,e,\sigma$. 
We  call such a state  a \emph{minefield state}.
Here $\sigma$  is  the  ``dynamic'' demand  on  the  expression
  $e$, i.e., it arises
  from the  actual sequence of function  calls that led
  to  the evaluation  of  $e$.  %% The  0-CFA demand  used
  %% during    liveness     analysis    is     a    static
  %% over-approximation   of   $\sigma$   that   considers
  %% \emph{all  possible calls}  that  could  lead to  the
  %% evaluation of  $e$.
  The  information in  a suspended
  evaluation  context  on  the   stack  $S$  is  also
  similarly augmented with its  demand.  Thus a stacked
  entry now takes  the form $(\rho, \ell,  e, \sigma)$.  %% A
  %% modification  of  the   small  step  semantics  which
  %% carries the extra information  is shown in the bottom
  %% table of Figure~\ref{fig:minefield-semantics}.
\item  $GC(\rho,  S,  \heap,  e, \sigma)$ models  a  liveness  based  garbage
  collection that returns $(\rho', S', \heap')$, where the changes in
  $\rho, S$ and  $\heap$ are due
  to certain references being replaced by $\bot$.  This represents the
  act of  not retaining  (effectively garbage collecting)  the objects
  pointed to by these references  during an actual garbage collection.
  This is done in the following way:
  \begin{enumerate} 
  \item \label{gc-1}Using the current  evaluation context $\rho, \stk,
    \heap,   e,  \sigma$   and   the  demand   transformers
    $\Lfonly$, we construct a liveness environment \Lv\ for the
    expression $e$.
      %%  If  $e$  is  an expression,  then  $  \Lv  =
    %% \mathcal{L}(e, \sigma, \Lfonly)$, else if $e$ is an
    %% application,  then  $  \Lv  =  ref(e,  \sigma,  \Lfonly)$.  
 We  similarly  define  the  liveness  environment  for  the
 expressions $e$ in each of the suspended evaluation contexts $(\rho, \ell,  e, \sigma)$ in $S$, giving
 a stack of liveness environments that we shall denote $\mathsf{SL}$.
  \item \emph{Garbage collection of objects pointed by root sets:} For
    the environment $\rho$ in the current  context and for each $x \in
    domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
    \Lv$.  Similarly,  for each  stack entry $(\rho,\ell,e,\sigma)$
    in $S$ 
    with  $\Lv'$   as  the   corresponding  liveness   environment  in
    $\mathsf{SL}$, and for each $x \in domain(\rho)$, we set $\rho(x)$
    to $\bot$ iff $x.\epsilon \notin \Lv'$.
  \item \emph{Garbage  collection of  objects in  the heap:}  For each
    location $\ell$,  we set $\heap(\ell)$  to $\bot$ iff there  is no
    variable  $x$  in either  the  current  environment or  a  stacked
    environment,  say  $\rho$,  such   that  $x.\alpha$  is  live  and
    corresponds  to   $\ell$ in \heap.  More  formally,  there   are  no  $x\in
    domain(\rho)$ and  $\alpha$, such that  $\heap\oxlb\rho(x), \alpha
    \oxrb = \ell$  and $x.\alpha \in \Lv$, where \Lv\  is the liveness
    environment corresponding to $\rho$ mentioned in step~\ref{gc-1}.
  \end{enumerate}
\end{enumerate}
Starting from the initial state,
%%  $([\;]_\rho,([\;]_\rho,
%% \ans, (\print~\ans)):[\;]_{S}  , [\;]_{H}, (\mainpgm)$,
every transition  is preceded  by a  garbage collection
using $GC$.  The details of the transition for the {\sc
  car-clo}  rule   is  shown   in
Figure~\ref{fig:minefield-semantics-for-some}.  If  this  or  an
earlier  call to  $GC$  results  in $H(\rho(x))$  being
bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
enters the  \bang\ state ({\sc  car-clo-bang}). Otherwise
the transition is the same as the earlier {\sc car-clo}
rule.

Consider a trace  of a minefield execution of a  program $p$, possibly
ending in  a \bang\ state. We  can construct a call-tree  based on the
trace in  which each node represents  a function that was  called (but
did not necessarily  return because of a \bang).  Assume  that each of
the nodes of  the tree is also annotated with  the program point where
the corresponding call  was invoked.  This tree can be  used to inline
function calls in a hierarchical fashion.  The details of the inlining
can be found in~\cite{asati14lgc}.
%% The details of inlining a single
%% function      call     $(\LET\,\,      x     \leftarrow
%% (f\,y_1\,\ldots\,y_n) \,\,  \IN\,\, e)$ is  as follows.
%% Let  $f$ be  defined  (after renaming  its formals  and
%% locals  to  be  disjoint from  existing  variables)  by
%% $(\DEFINE\ (f\  z_1\ \ldots\  z_n)\ e_f)$. The  call is
%% replaced  by   a  sequence  of  $\LET$s   of  the  form
%% $z_i\leftarrow (\ID\,\,y_i)$ followed by the body $e_f$
%% but with its  $(\SRETURN\,\,w)$ expressions replaced by
%% $(\LET\,\, x  \leftarrow (\ID\,\, w) \,\,  \IN\,\, e)$.
%% We prefer to use the  identity function $\ID$ as a form
%% of no-op rather than  introducing the form $(\LET\,\, x
%% \leftarrow  w \,\,  \IN\,\, e)$  where the  RHS of  the
%% $\LET$-definition   is  a   simple  variable. 
For  a
call-less program,  the initial state of  the minefield
semantics  is  assumed  to  be  $([\;]_\rho,([\;]_\rho,
\ans,      (\print~\ans)):[\;]_{S}     ,      [\;]_{H},
e_{\mainpgm}$.

@@

We  first  need  an auxiliary  result  about  minefield
semantics. Consider  a trace of a  minefield execution.
For every minefield state  $(\rho, S, H, e(s), \sigma)$
that appears  on the LHS of  a $\rightsquigarrow$ step,
the  demand   $\sigma$  on   the  expression   $e$  (or
application $s$) is non-null.  This can be proved by an
induction  on  the  number  of  steps  leading  to  the
minefield  state.   The  base step  holds  because  the
demand $\sigma_{all}$  on $(\mainpgm)$ is  non-null. For the 
inductive step  we observe   that for
each  step of  the minefield  semantics, if  the demand
$\sigma$ on  the LHS of  a minefield step  is non-null,
the demand on  the RHS is a  transformation of $\sigma$
(for example $\clazy\sigma$) which is also non-null.

@@

\begin{lemma}
\label{lemma:call-less-cannot-go-bang}
Consider the minefield execution of  a program without function calls.
Such a program cannot enter the \bang\ state.
\end{lemma}
\begin{proof}
Call a $GC(\ldots)$  followed by a $\rightsquigarrow$  transition as a
step. Consider  a state  $(\rho, S,  H, e,  \sigma)$ in  the minefield
execution of a  program.  We show by induction on  the number of steps
$n$ leading  to this  state that  the next  transition cannot  enter a
\bang\ state.  When $n$ is 0,  the state is $([\;]_\rho,\_, [\;]_{S} ,
            [\;]_{H}, e_{\mainpgm}, \sigma_{all})$.  Since the call to
            $GC(\ldots)$ in this state does  nothing, we just have to show
            that  the  $\rightsquigarrow$  transition cannot  enter  a
            \bang\   state.    Since   our  programs   are   in   ANF,
            $e_{\mainpgm}$ can  only be a $\LET$  expression.  A
                     {\sc let} step does not involve dereferencing,
                     and thus
                     cannot result in a \bang.

For the inductive step, we  shall show that none of the
minefield steps that  involves dereferencing results in
a   \bang.   These  are   the   steps   which  have   a
$H(\rho(...))$ in the premise. %%  Recall that a minefield
%% step   consists   of   a   GC  step   followed   by   a
%% $\rightsquigarrow$. 
Now a  $\rightsquigarrow$ step can  go
\bang\ because it dereferences a $\bot$ inserted by the
immediately  preceding  GC  or  the GC  of  an  earlier
minefield step.  However the  demand $\sigma'$
  on basis of which an earlier GC would have inserted a
  $\bot$,  would  have   included  the  current  demand
  $\sigma$.  Thus,  if the  earlier GC had  inserted a
$\bot$, so would the current  GC.  Thus it is enough to
consider the $\rightsquigarrow$  along with the current
GC and show that it does not lead to a \bang.

We consider the rules {\sc car-clo} and {\sc car-1-clo} only.  
The rest  of the rules involve similar reasoning. For the {\sc
  car-clo}  rule in  the state  $\rho, S,  H, (\CAR~x),
\sigma$, we  know that $\sigma$ is  non-null. Therefore
the  liveness  of  $x$  includes  $\epsilon$,  and  the
dereferencing $H(\rho(x))$ will go without \bang.

@@

\begin{lemma}
\label{lemma:call-less-cannot-go-bang}
Consider the minefield execution of  a program without function calls.
Such a program cannot enter the \bang\ state.
\end{lemma}
\begin{proof}
Consider a state in the minefield execution of a program $(\rho, S, H,
e, \sigma)$.   We show  by induction  on the  number $n$  of minefield
steps  leading to  this state  that  the next  step cannot  lead to  a
\bang\ state.

>>>>

%%           &\makecell{ $\rho, (\rho', x, e, \cred{\sigma'})\!:\!S,
%%   \heap, \kappa, \cred{\sigma}$  $\rightsquigarrow \rho', S, \heap[\rho'(x) :=
%%     \kappa], e, \cred{\sigma'}$ }   &  \sc{const}
%% \\
%% \hline
%%           & \makecell{$\rho, (\rho', z, e, \cred{\sigma'})\!:\!S, \heap, (\CONS~x~y), \cred{\sigma}$  $\rightsquigarrow
%%   \rho', S, \heap[\rho'(z) := (\rho(x),\rho(y))], e, \cred{\sigma'}$}     &  \sc{cons} \\
%% \hline
%% $\heap(\rho(x)) \mbox{ is } (\ell_1, \ell_2)$ & \makecell{$\rho, (\rho', z, e,
%%   \cred{\sigma'} )\!:\!S, \heap, (\CAR~x), \cred{\sigma}$  $
%%   \rightsquigarrow \rho', S, \heap[\rho'(z) := \heap(\ell_1)], e, \cred{\sigma'}$}      &
%% \sc{car-whnf} \\
%% \hline
%% $\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S,
%%   \heap, (\CAR~x), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, x,
%%   (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, \cred{(\clazy \cup \acar)\sigma }$}      &
%% \sc{car-clo}
%% \\

%% \hline
%% $\heap(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$ & $\rho,\, S,\,  \heap,\,
%% (\CAR~x), \cred{\sigma} \rightsquigarrow \rho', \,(\rho, addr(\langle
%% s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S,\, \heap,\, s, \, \cred{\sigma}$      &
%% \sc{car-1-clo} \\

%% \hline
%% $\heap(\rho(x)), \heap(\rho(y)) \in \mathbb{N}$
%%  & \makecell{$\rho, (\rho', z, e, \cred{\sigma'})\!:\!S, \heap,
%%   (+~x~y), \cred{\sigma}$   $\rightsquigarrow \rho', S, \heap[\rho'(z)
%%     := \heap(\rho'(x)) + \heap(\rho'(y))], e, \cred{\sigma'}$}      &
%% \sc{prim-whnf} \\
%% \hline
%% $\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S,
%%   \heap, (+~x~y), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, x,
%%   (+~x~y), \cred{\sigma})\!:\!S, \heap, s, \cred{\clazy\sigma}$}      &
%% \sc{prim-1-clo} \\
%% \hline
%% $\heap(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho,
%%   S, \heap, (+~x~y), \cred{\sigma}$  $\rightsquigarrow \rho', (\rho, y,
%%   (+~x~y), \cred{\sigma})\!:\!S, \heap, s, \cred{\clazy\sigma}$}      &
%% \sc{prim-2-clo} \\
%% \hline
%% {$\mathit{g}~\mbox{defined as}$
%% $~(\DEFINE~(g~\myvec{y})~e_{\mathit{g}})$}  & \makecell{$\rho, S, \heap,
%%   (g~\myvec{x}), \cred{\sigma}$  $\rightsquigarrow [\myvec{y} \mapsto
%%     \rho(\myvec{x})], S, \heap, e_{\mathit{g}}, \cred{\sigma}$}      &
%% \sc{funcall} \\
%% \hline
%% $\ell$ is a new location& \makecell{$\rho, S, \heap, (\LET~x\leftarrow
%%   s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, \heap[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
%%   \ell]\rangle], e, \cred{\sigma}$} &
%% \sc{let} \\ 
%% \hline
%% $\heap(\rho(x)) \ne 0$ & \makecell{$\rho, S, \heap, (\pi:\SIF~\psi:x~e_1~e_2),
%%   \cred{\sigma}$  $\rightsquigarrow \rho, S, \heap,  e_1, \cred{\sigma}$} & \sc{if-true} \\
%% \hline
%% $\heap(\rho(x)) = 0$ & \makecell{$\rho, S, \heap, (\pi:\SIF~\psi:x~e_1~e_2),
%%   \cred{\sigma}$   $\rightsquigarrow
%% \rho, S, \heap,  e_2, \cred{\sigma}$} & \sc{if-false} \\
%% \hline
%% $\heap(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, \heap,
%%   (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma}$ $\rightsquigarrow
%% \rho', (\rho, x, (\SIF~x~e_1~e_2),  \cred{\sigma})\!:\!S, \heap, s,
%% \cred{\clazy\sigma}$}
%% &
%% \sc{if-clo} \\
%% \hline
%% {$\heap(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho,
%%   (\rho', z, e, \cred{\sigma'})\!:\!S, \heap, (\SRETURN~x), \cred{\sigma}$  $\rightsquigarrow \rho', S, \heap[\rho'(z) := v], e,
%%   \cred{\sigma'}$} &
%% \sc{return-whnf}\\
%% \hline
%% $\heap(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, \heap,
%%   (\psi:\SRETURN~x), \cred{\sigma}$  $
%%   \rightsquigarrow$
%% $\rho',~ (\rho, x, (\SRETURN~x), \cred{\sigma})\!:\!S, \heap,  s,
%%   \cred{\sigma}$} &
%% \sc{return-clo} \\
%% \hline
%% \end{tabular}
%% \caption{Minefield semantics.\label{fig:minefield-semantics}}
%% \end{center}
%% \vskip -5mm
%% \end{figure*}

We shall now present a proof of soundness of the analysis presented in
Section~\ref{sec:liveness-analysis}.  %% \warning{While  the proof  is on
  %% the broad lines of  \cite{asati14lgc}, there are several differences
  %% due to the  lazy semantics of the language in  this paper.}
Here are
the ideas behind the proof.
\begin{enumerate}
\item       We   augment     the        standard       semantics
  (Figure~\ref{fig:lang-semantics}) to model  a liveness-based garbage
  collection  just  before  each  $\rightsquigarrow$  transition.  The
  simulated garbage collection starts from  the root-set and inserts a
  special value  $\bot$ at each location  in the heap that  contains a
  reference that is reachable but not live. Any attempt to dereference
  such  locations during  the  transition will  result  in entering  a
  special state denoted \bang. We call the semantics after
  augmentation,  \emph{minefield semantics}.
\item \label{inline1} Assuming that a  program enters the \bang\ state,
  we  show  how to  construct,  through  inline expansion,  a  program
  without function calls which has the same minefield behavior.
\item  The final  step   shows  that  no program  without
  function calls can enter a \bang\ state. As a consequence no program
  (with or without function calls) can enter the \bang\ state. 
\end{enumerate}

@@

\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S, \heap)}$,} $\heap(\rho(x)) \mbox{ is } \bot$ & $\rho, S,
  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$   &
\sc{car-clo-bang} 
\\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S, \heap)}$,} $\heap(\rho(x))
\mbox{ is } \langle s, \rho'\rangle$ & $\rho, S,
  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow   \rho', (\rho, x,
  (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, \cred{(\clazy \cup \acar)\sigma }$         &
\sc{car-clo}\\
\hline
\end{tabular}
\end{center}
\caption{Minefield semantics for \sc{car-clo-bang} and
  \sc{car-clo-bang} \label{fig:minefield-semantics-for-some}}
\end{figure*}

To set up the minefield semantics, we follow these steps:
\begin{enumerate}

\item   We   extend    the   abstract   machine   state
  $\rho,S,\heap,e$   to  $\rho,S,\heap,e,\sigma$. 
We  call such a state  a \emph{minefield state}.
Here $\sigma$  is  the  ``dynamic'' demand  on  the  expression
  $e$, i.e., it arises
  from the  actual sequence of function  calls that led
  to  the evaluation  of  $e$.  %% The  0-CFA demand  used
  %% during    liveness     analysis    is     a    static
  %% over-approximation   of   $\sigma$   that   considers
  %% \emph{all  possible calls}  that  could  lead to  the
  %% evaluation of  $e$.
  The  information in  a suspended
  evaluation  context  on  the   stack  $S$  is  also
  similarly augmented with its  demand.  Thus a stacked
  entry now takes  the form $(\rho, \ell,  e, \sigma)$.  %% A
  %% modification  of  the   small  step  semantics  which
  %% carries the extra information  is shown in the bottom
  %% table of Figure~\ref{fig:minefield-semantics}.
\item  $GC(\rho,  S,  \heap,  e, \sigma)$ models  a  liveness  based  garbage
  collection that returns $(\rho', S', \heap')$, where the changes in
  $\rho, S$ and  $\heap$ are due
  to certain references being replaced by $\bot$.  This represents the
  act of  not retaining  (effectively garbage collecting)  the objects
  pointed to by these references  during an actual garbage collection.
  This is done in the following way:
  \begin{enumerate} 
  \item \label{gc-1}Using the current  evaluation context $\rho, \stk,
    \heap,   e,  \sigma$   and   the  demand   transformers
    $\Lfonly$, we construct a liveness environment \Lv\ for the
    expression $e$.
      %%  If  $e$  is  an expression,  then  $  \Lv  =
    %% \mathcal{L}(e, \sigma, \Lfonly)$, else if $e$ is an
    %% application,  then  $  \Lv  =  ref(e,  \sigma,  \Lfonly)$.  
 We  similarly  define  the  liveness  environment  for  the
 expressions $e$ in each of the suspended evaluation contexts $(\rho, \ell,  e, \sigma)$ in $S$, giving
 a stack of liveness environments that we shall denote $\mathsf{SL}$.
  \item \emph{Garbage collection of objects pointed by root sets:} For
    the environment $\rho$ in the current  context and for each $x \in
    domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
    \Lv$.  Similarly,  for each  stack entry $(\rho,\ell,e,\sigma)$
    in $S$ 
    with  $\Lv'$   as  the   corresponding  liveness   environment  in
    $\mathsf{SL}$, and for each $x \in domain(\rho)$, we set $\rho(x)$
    to $\bot$ iff $x.\epsilon \notin \Lv'$.
  \item \emph{Garbage  collection of  objects in  the heap:}  For each
    location $\ell$,  we set $\heap(\ell)$  to $\bot$ iff there  is no
    variable  $x$  in either  the  current  environment or  a  stacked
    environment,  say  $\rho$,  such   that  $x.\alpha$  is  live  and
    corresponds  to   $\ell$ in \heap.  More  formally,  there   are  no  $x\in
    domain(\rho)$ and  $\alpha$, such that  $\heap\oxlb\rho(x), \alpha
    \oxrb = \ell$  and $x.\alpha \in \Lv$, where \Lv\  is the liveness
    environment corresponding to $\rho$ mentioned in step~\ref{gc-1}.
  \end{enumerate}
\end{enumerate}
Starting from the initial state,
%%  $([\;]_\rho,([\;]_\rho,
%% \ans, (\print~\ans)):[\;]_{S}  , [\;]_{\heap}, (\mainpgm)$,
every transition  is preceded  by a  garbage collection
using $GC$.  The details of the transition for the {\sc
  car-clo}  rule   is  shown   in
Figure~\ref{fig:minefield-semantics-for-some}.  If  this  or  an
earlier  call to  $GC$  results  in $\heap(\rho(x))$  being
bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
enters the  \bang\ state ({\sc  car-clo-bang}). Otherwise
the transition is the same as the earlier {\sc car-clo}
rule.

Consider a trace  of a minefield execution of a  program $p$, possibly
ending in  a \bang\ state. We  can construct a call-tree  based on the
trace in  which each node represents  a function that was  called (but
did not necessarily  return because of a \bang).  Assume  that each of
the nodes of  the tree is also annotated with  the program point where
the corresponding call  was invoked.  This tree can be  used to inline
function calls in a hierarchical fashion.  The details of the inlining
can be found in~\cite{asati14lgc}.
%% The details of inlining a single
%% function      call     $(\LET\,\,      x     \leftarrow
%% (f\,y_1\,\ldots\,y_n) \,\,  \IN\,\, e)$ is  as follows.
%% Let  $f$ be  defined  (after renaming  its formals  and
%% locals  to  be  disjoint from  existing  variables)  by
%% $(\DEFINE\ (f\  z_1\ \ldots\  z_n)\ e_f)$. The  call is
%% replaced  by   a  sequence  of  $\LET$s   of  the  form
%% $z_i\leftarrow (\ID\,\,y_i)$ followed by the body $e_f$
%% but with its  $(\SRETURN\,\,w)$ expressions replaced by
%% $(\LET\,\, x  \leftarrow (\ID\,\, w) \,\,  \IN\,\, e)$.
%% We prefer to use the  identity function $\ID$ as a form
%% of no-op rather than  introducing the form $(\LET\,\, x
%% \leftarrow  w \,\,  \IN\,\, e)$  where the  RHS of  the
%% $\LET$-definition   is  a   simple  variable. 
For  a
call-less program,  the initial state of  the minefield
semantics  is  assumed  to  be  $([\;]_\rho,([\;]_\rho,
\ans,      (\print~\ans)):[\;]_{S}     ,      [\;]_{\heap},
e_{\mainpgm}$.

@@

We  first  need  an auxiliary  result  about  minefield
semantics. Consider  a trace of a  minefield execution.
For every minefield state  $(\rho, S, \heap, e(s), \sigma)$
that appears  on the LHS of  a $\rightsquigarrow$ step,
the  demand   $\sigma$  on   the  expression   $e$  (or
application $s$) is non-null.  This can be proved by an
induction  on  the  number  of  steps  leading  to  the
minefield  state.   The  base step  holds  because  the
demand $\sigma_{all}$  on $(\mainpgm)$ is  non-null. For the 
inductive step  we observe   that for
each  step of  the minefield  semantics, if  the demand
$\sigma$ on  the LHS of  a minefield step  is non-null,
the demand on  the RHS is a  transformation of $\sigma$
(for example $\clazy\sigma$) which is also non-null.

@@

\begin{lemma}
\label{lemma:call-less-cannot-go-bang}
Consider the minefield execution of  a program without function calls.
Such a program cannot enter the \bang\ state.
\end{lemma}
\begin{proof}
Call a $GC(\ldots)$  followed by a $\rightsquigarrow$  transition as a
step. Consider  a state  $(\rho, S,  \heap, e,  \sigma)$ in  the minefield
execution of a  program.  We show by induction on  the number of steps
$n$ leading  to this  state that  the next  transition cannot  enter a
\bang\ state.  When $n$ is 0,  the state is $([\;]_\rho,\_, [\;]_{S} ,
            [\;]_{\heap}, e_{\mainpgm}, \sigma_{all})$.  Since the call to
            $GC(\ldots)$ in this state does  nothing, we just have to show
            that  the  $\rightsquigarrow$  transition cannot  enter  a
            \bang\   state.    Since   our  programs   are   in   ANF,
            $e_{\mainpgm}$ can  only be a $\LET$  expression.  A
                     {\sc let} step does not involve dereferencing,
                     and thus
                     cannot result in a \bang.

For the inductive step, we  shall show that none of the
minefield steps that  involves dereferencing results in
a   \bang.   These  are   the   steps   which  have   a
$\heap(\rho(...))$ in the premise. %%  Recall that a minefield
%% step   consists   of   a   GC  step   followed   by   a
%% $\rightsquigarrow$. 
Now a  $\rightsquigarrow$ step can  go
\bang\ because it dereferences a $\bot$ inserted by the
immediately  preceding  GC  or  the GC  of  an  earlier
minefield step.  However the  demand $\sigma'$
  on basis of which an earlier GC would have inserted a
  $\bot$,  would  have   included  the  current  demand
  $\sigma$.  Thus,  if the  earlier GC had  inserted a
$\bot$, so would the current  GC.  Thus it is enough to
consider the $\rightsquigarrow$  along with the current
GC and show that it does not lead to a \bang.

We consider the rules {\sc car-clo} and {\sc car-1-clo} only.  
The rest  of the rules involve similar reasoning. For the {\sc
  car-clo}  rule in  the state  $\rho, S,  \heap, (\CAR~x),
\sigma$, we  know that $\sigma$ is  non-null. Therefore
the  liveness  of  $x$  includes  $\epsilon$,  and  the
dereferencing $\heap(\rho(x))$ will go without \bang.

@@

\begin{lemma}
\label{lemma:call-less-cannot-go-bang-app}
Consider the minefield execution of  a program without function calls.
Such a program cannot enter the \bang\ state.
\end{lemma}
\begin{proof}
Consider a state in the minefield execution of a program $(\rho, S, \heap,
e, \sigma)$.   We show  by induction  on the  number $n$  of minefield
steps  leading to  this state  that  the next  step cannot  lead to  a
\bang\ state.

When $n$ is 0, the  state is $([\;]_\rho,\_, [\;]_{S} ,
[\;]_{\heap},  e_{\mainpgm},  \sigma_{all})$.  Further,  we
have   to  show   that   a  $GC(..)$   followed  by   a
$\rightsquigarrow$ step cannot lead  to a \bang\ state.
Now, because  our programs  are in  ANF, $e_{\mainpgm}$
can  only  be a  $\LET$.  A  {\sc  let} step  does  not
dereference anything and thus cannot result in a \bang.
